---
title: "Tuning and Pipelines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tuning and Pipelines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(tidylearn)
library(dplyr)
```

## Overview

`tl_model()` fits one model with the hyperparameters you name. Two families
build on that:

- **Tuning** searches a space of hyperparameters and returns the model fitted
  with the winner. `tl_tune_grid()` walks every combination; `tl_tune_random()`
  samples from ranges.
- **Pipelines** bundle preprocessing, a set of models, and an evaluation
  scheme into one object you can run, save, reload, and predict from.

The two compose: tune to find the settings, then put the winning settings in a
pipeline so the whole recipe is reproducible.

## Grid Search

`tl_tune_grid()` takes a named list of candidate values, one entry per
hyperparameter, and cross-validates every combination.

```{r}
tuned_tree <- tl_tune_grid(
  iris, Species ~ .,
  method = "tree",
  param_grid = list(cp = c(0.001, 0.01, 0.1), minsplit = c(5, 20)),
  folds = 3,
  verbose = FALSE
)
```

What comes back is an ordinary tidylearn model, already fitted with the
winning settings, so everything you would normally do with a model still
works:

```{r}
print(tuned_tree)
```

The search itself is attached as a `"tuning_results"` attribute:

```{r}
tuning <- attr(tuned_tree, "tuning_results")
names(tuning)
```

```{r}
tuning$results
```

```{r}
# The settings that won, and the score they won with
tuning$best_params
tuning$best_metric
```

The winning values reach the underlying fit, not just the report:

```{r}
tuned_tree$fit$control$cp
```

### Choosing the metric

Without a `metric`, tuning uses accuracy for classification and RMSE for
regression. Name one explicitly and the optimisation direction follows from
the metric: `rmse`, `mse`, `mae` and `mape` are minimised, everything else
maximised. Pass `maximize` only to override that.

```{r}
tuned_reg <- tl_tune_grid(
  mtcars, mpg ~ .,
  method = "forest",
  param_grid = list(mtry = c(2, 4), ntree = c(100, 300)),
  folds = 3,
  metric = "rmse",
  verbose = FALSE
)

attr(tuned_reg, "tuning_results")$best_params
```

### Starting from a default grid

`tl_default_param_grid()` supplies a reasonable starting grid per method, at
three sizes. Use it as a first pass, then narrow around whatever won.

```{r}
tl_default_param_grid("tree", size = "small")
```

```{r}
tl_default_param_grid("forest", size = "medium")
```

```{r}
tuned_default <- tl_tune_grid(
  iris, Species ~ .,
  method = "tree",
  param_grid = tl_default_param_grid("tree", size = "small"),
  folds = 3,
  verbose = FALSE
)

attr(tuned_default, "tuning_results")$best_params
```

## Random Search

Grid search cost is the product of the candidate counts, so it grows quickly.
`tl_tune_random()` samples `n_iter` points instead.

How `param_space` describes each parameter decides how it is sampled:

| Specification | Sampled as |
|---|---|
| Two numbers, `c(lo, hi)` | Uniform on the continuous interval |
| Three or more numbers | Drawn from exactly those values |
| A character vector | Drawn from those levels |
| A function of no arguments | Whatever the function returns |

A two-element vector is always continuous, so `minsplit = c(2, 40)` samples
values like 11.34. For a parameter that has to be a whole number, list the
candidates instead:

```{r}
tuned_random <- tl_tune_random(
  iris, Species ~ .,
  method = "tree",
  param_space = list(
    cp = c(0.0001, 0.2),                 # continuous
    minsplit = c(2, 5, 10, 20, 30, 40)   # drawn from these six
  ),
  n_iter = 8,
  folds = 3,
  seed = 42,
  verbose = FALSE
)

attr(tuned_random, "tuning_results")$best_params
```

Pass `seed` whenever you want the search to be reproducible. Without it, two
runs sample different points and can pick different winners.

## Looking at the Search

`tl_plot_tuning_results()` reads the attribute and draws it four ways.

```{r}
tl_plot_tuning_results(tuned_tree, plot_type = "scatter")
```

`"grid"` draws the two-parameter heat map that grid search is built for:

```{r}
tl_plot_tuning_results(tuned_tree, plot_type = "grid")
```

`"parallel"` puts every parameter on its own axis, which scales past two:

```{r}
tl_plot_tuning_results(tuned_tree, plot_type = "parallel")
```

`"importance"` ranks parameters by how much of the score variation each one
explains — a quick read on which knob is worth refining:

```{r}
tl_plot_tuning_results(tuned_tree, plot_type = "importance")
```

Every one of these is a ggplot2 object, so add to it as usual.

## Pipelines

A pipeline records preprocessing, the models to fit, and how to evaluate
them. Building it does no work; `tl_run_pipeline()` does.

```{r}
split <- tl_split(iris, prop = 0.7, stratify = "Species", seed = 42)

pipe <- tl_pipeline(
  split$train, Species ~ .,
  preprocessing = list(standardize = TRUE, dummy_encode = FALSE),
  models = list(
    tree = list(method = "tree"),
    forest = list(method = "forest", ntree = 300)
  ),
  evaluation = list(
    validation = "cv",
    cv_folds = 3,
    metrics = c("accuracy", "f1"),
    best_metric = "accuracy"
  )
)

print(pipe)
```

Anything you leave out of `preprocessing` or `evaluation` takes its default,
so a partial list is fine. An unrecognised name is an error rather than a step
that quietly does nothing.

```{r, error = TRUE}
tl_pipeline(split$train, Species ~ .,
            preprocessing = list(scale_method = "standardize"))
```

### Running it

```{r}
run <- tl_run_pipeline(pipe, verbose = FALSE)

names(run$models)
```

```{r}
print(run)
```

`tl_get_best_model()` returns the model that won on `best_metric`:

```{r}
best <- tl_get_best_model(run)
best$spec$method
```

### Predicting through the pipeline

This is the reason to use a pipeline rather than a bare model. Predicting on
raw new data replays the preprocessing the pipeline learned during the run,
applying the *training* centre and scale rather than recomputing them from
the new rows.

```{r}
preds <- tl_predict_pipeline(run, new_data = split$test, model_name = "forest")
head(preds)
```

```{r}
mean(preds$.pred == split$test$Species)
```

Omit `model_name` to predict with the best model.

### Saving and reloading

```{r}
path <- tempfile(fileext = ".rds")
tl_save_pipeline(run, path)

reloaded <- tl_load_pipeline(path)
names(reloaded$models)
```

```{r}
# Predictions survive the round trip, preprocessing included
reloaded_preds <- tl_predict_pipeline(
  reloaded, new_data = split$test, model_name = "forest"
)
identical(reloaded_preds$.pred, preds$.pred)
```

```{r, include = FALSE}
unlink(path)
```

## Tuning into a Pipeline

Tuning tells you the settings; the pipeline holds them alongside the
preprocessing that produced them.

```{r}
tuned <- tl_tune_grid(
  split$train, Species ~ .,
  method = "forest",
  param_grid = list(mtry = c(2, 3), ntree = c(100, 300)),
  folds = 3,
  verbose = FALSE
)

best_params <- attr(tuned, "tuning_results")$best_params
best_params
```

```{r}
final <- tl_pipeline(
  split$train, Species ~ .,
  models = list(
    forest = c(list(method = "forest"), best_params)
  ),
  evaluation = list(cv_folds = 3, metrics = "accuracy",
                    best_metric = "accuracy")
)

final_run <- tl_run_pipeline(final, verbose = FALSE)
final_preds <- tl_predict_pipeline(final_run, new_data = split$test)

mean(final_preds$.pred == split$test$Species)
```

## Cost

Tuning multiplies fits. A grid of *g* combinations at *k* folds is *g × k*
fits, plus one more to build the final model. The tree grid at the top of this
vignette is 6 combinations × 3 folds = 19 fits of a method that takes
milliseconds. The same grid on `method = "xgboost"` with 1000 rounds is the
same 19 fits of something much slower.

Two levers, in the order worth pulling:

1. **Fewer folds.** Going from 5 to 3 removes 40% of the work and still gives
   out-of-sample estimates.
2. **Random over grid.** `tl_tune_random(n_iter = 10)` costs a fixed 10 points
   regardless of how many parameters you are searching, where a grid over the
   same parameters costs their product.

`tl_compute_advisor()` estimates the cost of a single fit before you multiply
it by the search — worth a look before starting a long grid.

## Where to Go Next

- **AutoML** (`vignette("automl")`) searches across methods rather than within
  one, and manages its own budget.
- **Diagnostics** (`vignette("diagnostics")`) covers what to check once you
  have a fitted model.
